Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-18 main.tex
done with IX

\newcommand\aiD {\mathsf{ai}}
\newcommand\faD {\forall}
\newcommand\exD {\exists}
\newcommand\tttD {\ttt}
\renewcommand\wD {\mathsf{w}}
\newcommand\wlD {\mathsf{w}}
\newcommand\wrD {\mathsf{w}}
\renewcommand\cD {\mathsf{c}}
\renewcommand\acD {\mathsf{ac}}
\newcommand\acDeq {\mathsf{ac}^\fequ}
\newcommand\mfaD {\mathsf{m_\forall}}
\newcommand\mexD {\mathsf{m_\exists}}

@@

\newcommand{\Gr}{\mathcal{G}}
\newcommand{\PE}[1]{#1^\circ}

@@

In this section we establish the relation between ske bifibrations and derivations  in
$\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\mfaD$ and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\mfaD$ and
$\rectif\mexD$, show below:
\begin{equation*}
  \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
  \qquad
  \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
\end{equation*}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. In the rules $\rectif\mfaD$ and
$\rectif\mexD$ we also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

@@

\begin{lemma}\label{lem:cw->skew}
  Let $\vlderivation{
    \vlde{\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ be a recified derivation. Then the induced map
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ is a skew bifibration.
\end{lemma}

@@

\begin{proof}[Proof of Lemma~\ref{lem:cw->skew}]
  First, observe that for every inference rule
  $\rr\in\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we
  the induced map $\mapof{\rr}\colon\vgraphof Q\to\vgraphof P$ defines
  a fibration on the corresponding binding graphs. Therefore, their
  composition $\mapof\Deri$ is a fibration.

  Now we construct for
  $\Deri$ its propositional encoding $\PE\Deri$ by
  translating every line into its proposional encoding. The instances
  of the rules $\rectif\mfaD$ and $\rectif\mexD$ are replaced two
  steps by:
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlor(\PE{(Ax)}\vlor\PE{(Bx))}}}{
        \vlin{\fequ}{}{\Scons{(y\vlor z)\vlor(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlor\PE{(Ay)})\vlor (z\vlor\PE{(Bz)})}}}}}
  \end{equation*}
  and
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlan\PE{(Ax)}\vlor\PE{(Bx)}}}{
        \vlin{\me}{}{\Scons{(y\vlor z)\vlan(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlan\PE{(Ay)})\vlor (z\vlan\PE{(Bz)})}}}}}
  \end{equation*}
  respectively, where $\rectif\acD$ is a $\acD$ that renames the
  variabes---the atom variable, as well as the first-order variable of
  the same name---as everything is recified, there is no ambiguity
  here.
  This gives us a derivation $\vlderivation{
    \vlde{\set{\wrD,\acD,\rectif\acD,\me,\PE\fequ}}{\PE\Deri}{\PE
      B}{ \vlhy{\PE A}}}$ such that  $\mapof{\PE\Deri}=\mapof\Deri$. It has been
  shown in~\cite{str:07:RTA} that $\mapof{\PE\Deri}$ is a skew
  fibration (see
  also~\cite{hughes:invar,str:ral:tableaux19,str:RR-9048}). Hence,
  $\mapof\Deri$ is a skew fibration.
\end{proof}

@@

In the proof of this lemma, we make use of the following concept: Let
$\vlder{\sysS}{\DDeri}{Q}{P}$ be a derivation where $Q=\PE D$ is the
propostional encoding of some formula~$D$. We say that $\DDeri$ can be
\bfit{lifted} to $\sysS'$ if there is a formula $C$ such that
$P\fequ\PE C$ and there is a derivation
$\vlder{\sysS'}{\DDeri'}{D}{C}$.

\begin{proof}[Proof of Lemma~\ref{skew->cw}]
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Hence, we have two maps $\phi''\colon\gA\to\gB_1$ being a
  surjection and $\phi'\colon \gB_1\to\gB$ being an injection that
  reflects edges. Both, $\phi'$ and $\phi''$ remain skew
  bifibrations. Let us first look at $\phi'$.  Let $\PE B_1$ be the
  propositional formula obtained from $\PE B$ by removing all atoms
  that are not represented by vertices in $\vB'$. Then $\graphof{\PE
    B_1}=\gB_1$.  By \cite[Proposition~7.6.1]{str:07:RTA}, we have a
  derivation $\vlder{\set{\wrD,\fequ}}{\PE\Deri_1}{\PE B}{\PE B_1}$.
  A subformula of $\PE B$ is called \emph{weak} if it has been
  introduced by a weakening. All subformulas of a weak formula are
  also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  \emph{weak pair} if $B\fequ\Scons{B'\vlor B''}$ for some context
  $\Sconhole$. We now can assume without loss of generality that
  whenever weak subformulas $B'$ and $B''$ for a weak pair, they have
  been introduced by the same instance of $\wlD$ or $\wrD$ in
  $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  brought into this form by simple rule permutations, and then
  collapsing several weakenings into a single one.}  Now we show that
  $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  delete all atoms in the scope of the corresponding quantifier,
  because $\phi$ is a fibration on the binding graph. Hence, each line
  in $\PE\Deri_1$ is the propositional encoding $\PE P$ of a
  first-order formula $P$.  We now have to show
  that each instance of $\wrD$ is indeed a correct
  application in first-order logic. For this we have to inspect the
  cases a weakening happens inside a subformula $x\vlor C$ or $x\vlan
  C$ in $\PE\Deri_1$. There are the following cases:
  \begin{equation*}%\small
    \vlinf{}{}{S\cons{x\vlor D\vlor C}}{S\cons{x\vlor C}}
    \qquad
    \vlinf{}{}{S\cons{x\vlan (D\vlor C)}}{S\cons{x\vlan C}}
    \qquad
    \vlinf{}{}{S\cons{(x\vlor D)\vlan C}}{S\cons{x\vlan C}}
  \end{equation*}
  In the first case the weakening happens inside the scope of a
  $\forall$-quantifier, and in the second case inside the scope of a
  $\exists$-quantifier. Both cases are unproblematic in first-order
  logic. However, in the third case an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur.
  Thus we have a first order derivation
  $\vlder{\set{\wrD,\fequ}}{\Deri_1}{B}{B_1}$.

@@

  where $\PE A_1 =\PE A\substof\phi$ is the result of applying $\substof\phi$
  to~$\PE A$. Note that $\PE A_1 = \PE{(A\substof\phi)}$. To $\Deri_2$ we can apply a permutation argument,
  permuting all instances of $\acD$, that duplicate an atom $x\in\VAR$, up until they either reach the top
  of the derivation or an instance of $\me$ which separates the two
  atoms in the premise. More precisely, we consider the following
  inferene rule

@@

@@

  Now we show that $\PE\Deri_2$ can be lifted to
  $\set{\acD,\me,\mfaD,\mexD,\fequ}$.  We proceed by induction on the
  height of $\PE\Deri_2$, beginning at the bottom, as we already
  constructed the formula~$B_1$.  Every instance of $\fequ$ or $\acD$
  (duplicating an atom that is not a $x\in\VAR$), can trivially be
  lifted. Similarly, an instance of $\me$, that is above an instance
  of $\acDeq$ as indicated in~\eqref{eq:mac}, can be lifted as the
  premise must also be a proposional encoding and the $\me$-inference
  remains correct in the first-order case.  It remains to consider a
  situation~\eqref{eq:mac}. Observe that we must have

@@

    previous one.

@@

  Finally, we need to consider the instances of $\acDeq$ that are at the top of the derivation.

>>>>

\newcommand\aiD {\mathsf{ai}}
\newcommand\faD {\forall}
\newcommand\exD {\exists}
\newcommand\tttD {\ttt}
\renewcommand\wD {\mathsf{w}}
\newcommand\wlD {\mathsf{w}}
\newcommand\wrD {\mathsf{w}}
\renewcommand\cD {\mathsf{c}}
\renewcommand\acD {\mathsf{ac}}
\newcommand\acDx {\mathsf{ac}_x}
\newcommand\acDeq {\mathsf{ac}_x^\fequ}
\newcommand\cfaD {\mathsf{c_\forall}}
\newcommand\mfaD {\mathsf{m_\forall}}
\newcommand\mexD {\mathsf{m_\exists}}

@@

\newcommand{\Gr}{\mathcal{G}}
\newcommand{\PE}[1]{#1^\circ}
\newcommand{\PEp}[1]{\tilde{#1}}

@@

\todo{add $\cfaD$ everywhere}

In this section we establish the relation between ske bifibrations and derivations  in
$\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\mfaD$ and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\mfaD$ and
$\rectif\mexD$, show below:
\begin{equation*}
  \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
  \qquad
  \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
\end{equation*}
\todo{same for $\cfaD$}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. In the rules $\rectif\mfaD$ and
$\rectif\mexD$ we also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

@@

\begin{lemma}\label{lem:cw->skew}
  Let $\vlderivation{
    \vlde{\set{\wrD,\acD,\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ be a recified derivation. Then the induced map
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ is a skew bifibration.
\end{lemma}

@@

\begin{proof}[Proof of Lemma~\ref{lem:cw->skew}]
  First, observe that for every inference rule
  $\rr\in\set{\wrD,\acD,\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we
  the induced map $\mapof{\rr}\colon\vgraphof Q\to\vgraphof P$ defines
  a fibration on the corresponding binding graphs. Therefore, their
  composition $\mapof\Deri$ is a fibration.

  Now we construct for
  $\Deri$ its propositional encoding $\PE\Deri$ by
  translating every line into its proposional encoding. The instances
  of the rules $\rectif\mfaD$ and $\rectif\mexD$ are replaced two
  steps by:
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlor(\PE{(Ax)}\vlor\PE{(Bx)})}}{
        \vlin{\fequ}{}{\Scons{(y\vlor z)\vlor(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlor\PE{(Ay)})\vlor (z\vlor\PE{(Bz)})}}}}}
  \end{equation*}
  and
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlan(\PE{(Ax)}\vlor\PE{(Bx)})}}{
        \vlin{\me}{}{\Scons{(y\vlor z)\vlan(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlan\PE{(Ay)})\vlor (z\vlan\PE{(Bz)})}}}}}
  \end{equation*}
  respectively, where $\rectif\acD$ is a $\acD$ that renames the
  variabes---the atom variable, as well as the first-order variable of
  the same name---as everything is recified, there is no ambiguity
  here.  This gives us a derivation $\vlderivation{
    \vlde{\set{\wrD,\acD,\rectif\acD,\me,\PE\fequ}}{\PE\Deri}{\PE B}{
      \vlhy{\PE A}}}$ such that $\mapof{\PE\Deri}=\mapof\Deri$. It has
  been shown in~\cite{str:07:RTA} that $\mapof{\PE\Deri}$ is a skew
  fibration (see
  also~\cite{hughes:invar,str:ral:tableaux19,str:RR-9048}). Hence,
  $\mapof\Deri$ is a skew fibration. That $\mapof{\PE\Deri}$ preserves
  exponentials follows immediately from its construction.
\end{proof}

@@

In the proof of this lemma, we make use of the following concept: Let
$\vlder{\sysS}{\DDeri}{Q}{P}$ be a derivation where $P$ and $Q$ are
propositional formulas (possibly using variable $x\in\VAR$ at the
places of atoms).  We say that $\DDeri$ can be \bfit{lifted} to
$\sysS'$ if there are (first-order) formulas $C$ and $D$ such that $P=\PE C$ and
$Q=\PE D$ and there is a derivation $\vlder{\sysS'}{\DDeri'}{D}{C}$.

\begin{proof}[Proof of Lemma~\ref{skew->cw}]
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Hence, we have two maps $\phi''\colon\gA\to\gB_1$ being a
  surjection and $\phi'\colon \gB_1\to\gB$ being an injection that
  reflects edges. Both, $\phi'$ and $\phi''$ remain skew
  bifibrations. Let us first look at $\phi'$.  Let $\PEp B_1$ be the
  propositional formula obtained from $\PE B$ by removing all atoms
  that are not represented by vertices in $\vB'$. Then $\graphof{\PEp
    B_1}=\gB_1$.  By \cite[Proposition~7.6.1]{str:07:RTA}, we have a
  derivation $\vlder{\set{\wrD,\fequ}}{\PE\Deri_1}{\PE B}{\PEp B_1}$.
  A subformula of $\PE B$ is called \emph{weak} if it has been
  introduced by a weakening. All subformulas of a weak formula are
  also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  \emph{weak pair} if $B\fequ\Scons{B'\vlor B''}$ for some context
  $\Sconhole$. We now can assume without loss of generality that
  whenever weak subformulas $B'$ and $B''$ for a weak pair, they have
  been introduced by the same instance of $\wlD$ or $\wrD$ in
  $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  brought into this form by simple rule permutations, and then
  collapsing several weakenings into a single one.}  Now we show that
  $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  delete all atoms in the scope of the corresponding quantifier,
  because $\phi$ is a fibration on the binding graph. Hence, each line
  in $\PE\Deri_1$ is the propositional encoding $\PE P$ of a
  first-order formula $P$.  We now have to show
  that each instance of $\wrD$ is indeed a correct
  application in first-order logic. For this we have to inspect the
  cases a weakening happens inside a subformula $x\vlor C$ or $x\vlan
  C$ in $\PE\Deri_1$. There are the following cases:
  \begin{equation*}%\small
    \vlinf{}{}{S\cons{x\vlor D\vlor C}}{S\cons{x\vlor C}}
    \qquad
    \vlinf{}{}{S\cons{x\vlan (D\vlor C)}}{S\cons{x\vlan C}}
    \qquad
    \vlinf{}{}{S\cons{(x\vlor D)\vlan C}}{S\cons{x\vlan C}}
  \end{equation*}
  In the first case the weakening happens inside the scope of a
  $\forall$-quantifier, and in the second case inside the scope of a
  $\exists$-quantifier. Both cases are unproblematic in first-order
  logic. However, in the third case an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur.
  Thus we have a first order derivation
  $\vlder{\set{\wrD,\fequ}}{\Deri_1}{B}{B_1}$ with $\PE B_1=\PEp B_1$.

@@

  where $\PE A_1 =\PE A\substof\phi$ is the result of applying
  $\substof\phi$ to~$\PE A$. Note that $\PE A_1 =
  \PE{(A\substof\phi)}$ and $\PE B_1$ are both proposional
  encodings. We plan to show that $\Deri_2$ can be lifted to
  $\set{\acD,\cfaD,\me,\mfaD,\mexD,\fequ}$. However, observe that not
  every formula occurring in $\Deri_2$ is a propositional
  encoding. There are two reasons for this: (i) we might have $P\PE\fequ
  Q$ where $P$ is a propositional encoding but $Q$ is not, and (ii)
  the rule $\acD$ can duplicate an atom with name $x\in\VAR$. Let us write $\acDx$ for such instances. To
  address (i), we consider here formulas equivalent modulo $\fequ$,
  always knowing that we can add instances of $\fequ$ as needed. To
  address (ii), we apply a permutation argument, permuting all
  instances of $\acDx$ up until
  they either reach the top of the derivation or an instance of $\me$
  which separates the two atoms in the premise. More precisely, we
  consider the following inferene rule

@@

    \label{eq:acDeq}

@@

  We can therefore assume that all instances of $\acDx$, that contract
  an atom $x\in\VAR$ are either at the top of $\PE\Deri_2$ or below a
  $\me$-instance as in~\eqref{eq:mac}. We now lift $\PE\Deri_2$ to
  $\set{\acD,\cfaD,\me,\mfaD,\mexD,\fequ}$, proceed by induction on
  the height of $\PE\Deri_2$, beginning at the top, making a case
  analysis on the topmost rule that is not a $\fequ$.
  \item $\acDx$: We know that the premisse of~\eqref{eq:acDeq} is a
    propositional encoding. Hence, $S_1\conhole=\conhole\vlor E$ and
    $S_2\conhole=\conhole\vlor F$ and both $x$ are universals, and
    $E\vlor F$ contains all occurrences of $x$ bound by that
    universal. We have the following subcases:
    \item $E$ and $F$ are both non-empty: We have
        \vlinf{\acDeq}{}{\Scons{x\vlor (E\vlor F)}}{S\cons{(x\vlor E)\vlor (x\vlor F)}}
        \vlinf{\mfaD}{}{\Scons{\forall x.(E\vlor F)}}{S\cons{(\forall x.E)\vlor (\forall x.F)}}
    \item $E$ is empty and $F$ is non-empty: We have
        \vlinf{\acDeq}{}{\Scons{x\vlor F)}}{S\cons{x\vlor (x\vlor F)}}
        \vlinf{\mfaD}{}{\Scons{\forall x.F}}{S\cons{\forall x.\forall x.F}}
    \item $E$ is non-empty and $F$ is empty: This is similar to the previous case.
    \item $E$ and $F$ are both empty: This is impossible as the
      premise would not be a propositional encoding.
  \item $\acD$ (contracting an ordinary atom): This can trivially be lifted.
  \item $\me$ that is not in the situation of~\eqref{eq:mac}: Then now
    encoding of a quantifier is affected and the instance of $\me$ can
    be lifted.
  \item $\me/\acDx$ as in situation~\eqref{eq:mac}: We must have

@@

      previous subcase.

@@

  Thus $\PE\Deri_2$ can be lifted to
  $\vlder{\set{\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{B_1}{A\substof\phi}$. We
  construct $\Deri$ by composing $\Deri_2$ and $\Deri_1$. Then
  $\rectif\Deri$ can be constructed by rectifying $\Deri$, where the
  variables to be used in $A$ are already given. That
  $\phi=\mapof{\rectif\Deri}$ follows immediately from the
  construction.
